---
title: "scRNA-seq Analysis with Seurat"
author: "Theo Killian"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
---

```{r, echo = FALSE, results = 'asis'}
suppressPackageStartupMessages(library("dplyr"))
knitr::opts_chunk$set(collapse=TRUE, comment="#>", warning=FALSE, message=FALSE)
```

# Introduction

This document demonstrates how to load and preprocess single-cell RNA-seq data
generated via 10x Genomics (and processed by Cell Ranger) using the Seurat R package. We will:

1) Read the filtered feature-barcode matrix (barcodes.tsv.gz, features.tsv.gz, and matrix.mtx.gz) into R.
1) Create a Seurat object.
1) Perform preliminary QC and filtering (removing low-quality cells and potential artifacts).
1) Run dimensionality reduction (PCA, UMAP), clustering, and visualize the results.

```{r load_libraries, warning=FALSE, message=FALSE}
library(dplyr)
library(tidyr)
library(tibble)
library(stringr)
library(readxl)
library(writexl)
library(Seurat)       # For single-cell data analysis
library(ggplot2)      # Visualization
library(patchwork)    # Combine multiple ggplot2 plots
```

## Load 10x Genomics Data

Below, we assume that your filtered matrix files are in a directory like:

```{bash}
results/cellranger_count/SAMPLE_1/outs/filtered_feature_bc_matrix/
```

Adjust this path for your dataset as needed.

```{r warning=FALSE, message=FALSE}
# Replace "SAMPLE_1" with your actual sample directory
data_dir <- "results/cellranger_count/SAMPLE_1/outs/filtered_feature_bc_matrix/"

# Read10X() automatically looks for barcodes.tsv.gz, features.tsv.gz, matrix.mtx.gz
# in the provided directory
sc_data <- Read10X(data.dir = data_dir)

# Create a Seurat object
# min.cells = 3 means a gene is retained if detected in at least 3 cells
# min.features = 200 means a cell is retained if it has at least 200 genes detected
seurat_obj <- CreateSeuratObject(
  counts = sc_data,
  project = "MyProject",
  min.cells = 3,
  min.features = 200
)

# Basic exploration
seurat_obj
```

## Quality Control (QC)

Cells with low unique feature counts or high mitochondrial gene percentages are
often considered low-quality or dying cells. We can identify them by calculating
the percentage of reads mapped to mitochondrial genes and then filtering.

```{r warning=FALSE, message=FALSE}
# Calculate percent of transcripts mapping to mitochondrial genes
# Note: Human mitochondrial genes often start with "MT-"; adjust for other species
seurat_obj[["percent.mt"]] <- PercentageFeatureSet(seurat_obj, pattern = "^MT-")

# Visualize QC metrics
VlnPlot(seurat_obj, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)

# Scatter plot to explore relationships between metrics
FeatureScatter(seurat_obj, feature1 = "nCount_RNA", feature2 = "percent.mt") +
  FeatureScatter(seurat_obj, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")

# Basic filtering thresholds:
# - Keep cells with 200 < nFeature_RNA < 2500
# - Keep cells with < 5% mitochondria
seurat_obj_filtered <- subset(
  seurat_obj,
  subset = nFeature_RNA > 200 & nFeature_RNA < 2500 & percent.mt < 5
)

# Check the updated object
seurat_obj_filtered
```

## Normalize Data and Identify Variable Features

Next, we normalize the data to account for differences in library size, then identify the most variable genes.

```{r warning=FALSE, message=FALSE}
# Log-normalize the data
seurat_obj_filtered <- NormalizeData(seurat_obj_filtered)

# Find the most variable features
seurat_obj_filtered <- FindVariableFeatures(
  seurat_obj_filtered,
  selection.method = "vst",
  nfeatures = 2000
)

# Inspect the top 10 most variable genes
top10 <- head(VariableFeatures(seurat_obj_filtered), 10)
top10
```

We can also visualize the variable features:

```{r fig.width=10, warning=FALSE, message=FALSE}
plot1 <- VariableFeaturePlot(seurat_obj_filtered)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot1 + plot2
```

## Scaling and Linear Dimensional Reduction (PCA)

```{r fig.width=10, warning=FALSE, message=FALSE}
# Scale the data
seurat_obj_filtered <- ScaleData(seurat_obj_filtered)

# Run PCA
seurat_obj_filtered <- RunPCA(seurat_obj_filtered, features = VariableFeatures(seurat_obj_filtered))

# Visualize the PCA
DimPlot(seurat_obj_filtered, reduction = "pca")

# Examine PCA results
ElbowPlot(seurat_obj_filtered)
```

The ElbowPlot helps decide how many principal components (PCs) to use for downstream clustering.

## Clustering and UMAP

We typically choose a range of PCs (e.g., 1:10 or 1:15) to capture the major
sources of variation. Then we run clustering and UMAP:

```{r warning=FALSE, message=FALSE}
# Find neighbors using the first 10 PCs (adjust as needed)
seurat_obj_filtered <- FindNeighbors(seurat_obj_filtered, dims = 1:10)

# Find clusters at a chosen resolution (adjust resolution as needed for your dataset)
seurat_obj_filtered <- FindClusters(seurat_obj_filtered, resolution = 0.5)

# Run UMAP
seurat_obj_filtered <- RunUMAP(seurat_obj_filtered, dims = 1:10)

# UMAP visualization
DimPlot(seurat_obj_filtered, reduction = "umap", label = TRUE, pt.size = 0.6) +
  ggtitle("UMAP Clusters") + theme_minimal()
```

## Marker Gene Identification (Optional)

To identify marker genes for each cluster, run FindAllMarkers or FindMarkers for
specific clusters. This step can be computationally intensive:

```{r warning=FALSE, message=FALSE}
cluster_markers <- FindAllMarkers(
  seurat_obj_filtered, 
  only.pos = TRUE,
  min.pct = 0.25,
  logfc.threshold = 0.25
)
head(cluster_markers)
```

## Summary and Next Steps

We have now:

1) Loaded single-cell RNA-seq data from the Cell Ranger output directory.
1) Created a Seurat object and performed QC.
1) Filtered out low-quality cells.
1) Normalized data and identified highly variable genes.
1)Scaled data, ran PCA, performed clustering, and visualized via UMAP.

Next steps could include:

1) Annotating clusters by comparing with known marker genes.
1) Exploring sub-clusters or running trajectory analysis (if applicable).
1) Integrating multiple samples or time points (using Seurat’s integration workflows).
1) Further QC (e.g., doublet detection) if needed.

## Save results

The results are saved as an RDS object.

```{r warning=FALSE, message=FALSE}
saveRDS(object = seurat_obj_filtered, file = paste0(
  "path/to/file/seurat_obj_filtered_", Sys.Date(), ".rds"))
```

*Session Info*

```{r sessionInfo}
sessionInfo()
```
